Disturbance-aware minimum-time planning strategies for motorsport vehicles with probabilistic safety certificates

### 2025-05-05 intro.tex
[MGa] added discussion on LLE
and why they are of limited use in this context

In several related fields, various strategies have been developed to address planning under uncertainty.

\textit{Orbital Mechanics.} In orbital mechanics, uncertainty in the state estimate necessitates a probabilistic treatment of collision events between satellites and space debris. A foundational reference in this context is~\cite{Tapley:StatisticalOrbitDetermination:2004}, where statistical orbit determination techniques are employed to assess and mitigate collision risks.

\textit{Robotics.} In robotic motion planning, uncertainty-aware techniques are crucial to ensure safe navigation in environments populated with obstacles. One of the earliest contributions proposing a probabilistic representation of uncertainty is the chance-constrained framework in~\cite{Blackmore:ChanceConstrainedOptimalPath:2011}, which plans over the predicted probabilistic distribution of the system state to ensure that the probability of constraint violation remains below a specified threshold.

\textit{Model Predictive Control (MPC).} Within the MPC paradigm, recent works have incorporated probabilistic safety guarantees. Notably, the methods presented in~\cite{Gao:CollisionfreeMotionPlanning:2023},~\cite{Zhang:RobustifiedTimeoptimalPointtopoint:2025}, and~\cite{Zhang:RobustifiedTimeoptimalCollisionfree:2024} introduce stochastic MPC frameworks for autonomous mobile platforms, where process noise is explicitly modeled and closed-loop tracking performance is maintained via either pre-computed or optimized feedback gains. The approach in~\cite{Gao:CollisionfreeMotionPlanning:2023} additionally proposes a zero-order optimization scheme to include the feedback gain directly in the optimal control problem, mitigating the growth of uncertainty over the planning horizon.

\textit{Chemical Process Control.} In the domain of chemical engineering, robust MPC approaches have been proposed to address uncertainty in industrial settings. For example,~\cite{Krog:SimpleFastRobust:2024} presents a heuristic method based on $n$-step-ahead uncertainty predictions, which are used to compute constraint tightening margins for the control of a polymerization reactor.

\textit{Planning with Parametric Uncertainty.} In some contexts, uncertainty arises from partial or imprecise knowledge of system parameters. This has led to a significant body of work on robust trajectory planning, particularly for unmanned aerial vehicles (UAVs), with the aim of minimizing sensitivity to state and input variations. For instance,~\cite{Brault:RobustTrajectoryPlanning:2021},~\cite{Giordano:TrajectoryGenerationMinimum:2018}, and~\cite{Brault:TubebasedTrajectoryOptimization:} propose tube-based and sensitivity-aware optimization frameworks that explicitly account for both input and closed-loop state sensitivity in the planning phase. Further developments integrate observability-aware planning into a unified multi-step optimization framework, as demonstrated in~\cite{Bohm:COPControlObservabilityaware:2022}.

>>>>

In several science fields, different strategies have been developed to address planning under uncertainty.
In orbital mechanics, uncertainty in the state estimate necessitates a probabilistic treatment of collision events between satellites and space debris. A foundational reference in this context is~\cite{Tapley:StatisticalOrbitDetermination:2004}, where statistical orbit determination techniques are employed to assess and mitigate collision risks.

In robotic motion planning, uncertainty-aware techniques are crucial to ensure safe navigation in environments populated with obstacles. One of the earliest contributions proposing a probabilistic representation of uncertainty is the chance-constrained framework in~\cite{Blackmore:ChanceConstrainedOptimalPath:2011}, which plans over the predicted probabilistic distribution of the system state to ensure that the probability of constraint violation remains below a specified threshold.

Within the Model Predictive Control paradigm, recent works have incorporated probabilistic safety guarantees. Notably, the methods presented in~\cite{Gao:CollisionfreeMotionPlanning:2023},~\cite{Zhang:RobustifiedTimeoptimalPointtopoint:2025}, and~\cite{Zhang:RobustifiedTimeoptimalCollisionfree:2024} introduce stochastic MPC frameworks for autonomous mobile platforms, where process noise is explicitly modeled and closed-loop tracking performance is maintained via either pre-computed or optimized feedback gains. The approach in~\cite{Gao:CollisionfreeMotionPlanning:2023} additionally proposes a zero-order optimization scheme to include the feedback gain directly in the optimal control problem, mitigating the growth of uncertainty over the planning horizon.

In the domain of chemical engineering, robust MPC approaches have been proposed to address uncertainty in industrial settings. For example,~\cite{Krog:SimpleFastRobust:2024} presents a heuristic method based on $n$-step-ahead uncertainty predictions, which are used to compute constraint tightening margins for the control of a polymerization reactor.

In some contexts, uncertainty arises from partial or imprecise knowledge of system parameters. This has led to a significant body of work on robust trajectory planning, particularly for unmanned aerial vehicles (UAVs), with the aim of minimizing sensitivity to state and input variations. For instance,~\cite{Brault:RobustTrajectoryPlanning:2021},~\cite{Giordano:TrajectoryGenerationMinimum:2018}, and~\cite{Brault:TubebasedTrajectoryOptimization:} propose tube-based and sensitivity-aware optimization frameworks that explicitly account for both input and closed-loop state sensitivity in the planning phase. Further developments integrate observability-aware planning into a unified multi-step optimization framework, as demonstrated in~\cite{Bohm:COPControlObservabilityaware:2022}.

%In other approaches, the largest Lyapunov exponent (LLE) is proposed as an indicator of local stability~\cite{Meng:AnalysisGlobalCharacteristics:2022}. Lyapunov exponents, equal in number to the system's dimension, characterize the average exponential rate at which nearby trajectories of a nonlinear dynamical system converge or diverge~\cite{Strogatz:NonlinearDynamicsChaos:2019}. With the exception of chaotic systems - where a positive LLE may still be associated with bounded motion on a strange attractor - a negative LLE indicates local stability, as nearby trajectories tend to converge.
%
%A notable property is that, for trajectories approaching a stable equilibrium, the largest Lyapunov exponent asymptotically converges to the maximum real part of the eigenvalues of the linearized system at the equilibrium point.

Following a different route, the largest Lyapunov exponent (LLE) has been proposed as an indicator of local stability~\cite{Meng:AnalysisGlobalCharacteristics:2022}, as it quantifies the average exponential rate at which nearby trajectories of a nonlinear system converge (negative) or diverge (positive)~\cite{Strogatz:NonlinearDynamicsChaos:2019}. Except in chaotic systems where a positive LLE may still correspond to bounded motion on a strange attractor - a negative LLE typically signals local stability. Notably, for trajectories converging to a stable equilibrium, the LLE asymptotically approaches the largest real part of the eigenvalues of the linearized system at that point.

This result has been leveraged in the literature to infer whether a system, not yet at equilibrium, is ultimately converging toward one, thereby enabling a localized assessment of stability based on the transient behavior~\cite{Sadri:StabilityAnalysisNonlinear:2013}. Applications of this method are particularly relevant in domains such as chaotic vessel motions~\cite{McCue:UseLyapunovExponents:2011} and rotorcraft dynamics~\cite{Tamer:StabilityNonlinearTimeDependent:2016,Cassoni:RotorcraftStabilityAnalysis:2024}, where the primary concern is understanding the long-term evolution of the system and determining its asymptotic behavior.

%However, while Lyapunov-based indicators have been applied in vehicular contexts as well~\cite{Sadri:StabilityAnalysisNonlinear:2013,Meng:AnalysisGlobalCharacteristics:2022}, their relevance here is arguably of reduced interest - not due to their binary nature, but because even identifying asymptotic stability regions offers limited practical insight during planning. A trajectory may lie within the basin of attraction of a distant equilibrium, yet the long-term convergence may be of little operational use. What matters instead is whether, in the short term, a realistically modeled driver - acting in feedback or planning a bit into the future - can ensure that the vehicle's response to perturbations remains within safety-critical constraints, such as grip limits and collision avoidance. This perspective prioritizes short-horizon constraint-satisfying stabilizability over asymptotic notions of stability.

Although Lyapunov-based indicators have been applied in vehicular contexts~\cite{Sadri:StabilityAnalysisNonlinear:2013,Meng:AnalysisGlobalCharacteristics:2022}, their utility in planning is limited by the fact that asymptotic stability - while theoretically informative - often lacks operational relevance. A point along the planned trajectory may belong to the basin of attraction of a stable equilibrium, but if that equilibrium lies far ahead in time or outside the operational domain (e.g., off track), its practical relevance is limited.

%A point along a planned trajectory may lie within the basin of attraction of a stable equilibrium, yet if that equilibrium is reached only asymptotically, possibly over a long horizon or outside the intended operational domain (e.g., off track), its practical utility diminishes. 
In contrast, what is of primary interest in our setting is whether perturbations, even in transiently unstable conditions, can be effectively managed -- either through feedback or short-horizon planning -- so that the system remains within safety-critical constraints such as tire force limits and collision avoidance. This shifts the focus from long-term convergence toward evaluating whether the planned trajectory permits short-term stabilizability under uncertainty while ensuring constraint satisfaction.
